\vspace{-2cm} \LARGE Divergence-Free Smoothed Particle Hydrodynamics \\ \large Advanced Simulation and Visualization of Fluids in Computer Graphics

### 2016-12-14 report/paper.tex
divergencefree solver

\subsection{Neighbourhood search}
    Since SPH only considers a finite amount of neighbouring particles it is important to keep track of every particles neighbours.
    Searching through all particles for neighbours within the cutoff distance $h$ for every particle is inefficient and takes $\mathcal{O}({N^2})$ time.
    The cutoff distance $h$ is the kernels smoothing radius.
    To make this faster a cell list was implemented.
    A cell list is a data structure that is divided into cells that have a length larger or equals to the cutoff distance $h$.
    Each cell spans on a surface in space.
    A particle belongs to a cell if it is inside the space the cell occupy.
    When finding the neighbour of particle $i$, only the neighbouring cells have to be searched for particles within the cutoff distance, see figure~\ref{fig:cellList}.

    \begin{figure}[h!]\centering
    \begin{centering}
    \includegraphics[width=0.7\linewidth]{img/CellList.png}
    \caption{Finding the neighbours for the filled in particle $i$ by looking through all neighbouring cells, including its own cell, for particles within the cutoff distance $h$}
    \label{fig:cellList}
    \end{centering}
    \end{figure}

    The cell list is implemented by using a four dimensional vector where the first three dimensions are for the x, y and z coordinates and the fourth dimension is for storing the particles belonging to the cell.
    The amount of cells are decided by dividing the scene into cells of length $h$.
    The particles are then assigned to a cell by finding the cell's coordinates that it belongs to according to Equation ~\ref{eq:assignCell}, where $\vect{x}$ is the position of the particle and $\vect{s}_min$ is the lowest position of the scene space.
    If the particle moves out of its cell it is reassigned to the new cell it is inside.

    \begin{equation} \label{eq:assignCell}
        \frac{\lceil{\vect{x} - \vect{s}_{min}}\rceil}{h}
    \end{equation}

@@

    \begin{equation} \label{eq:kernel}
        \resizebox{0.88\linewidth}{!}{$
        W_h(q(x)) =  \frac{1}{\pi h^3} \left\{\begin{matrix}
        1 - \frac{3}{2}q^2 + \frac{3}{4}q^3 & 0 \leqslant q < 1 \\ 
        \\
        \frac{1}{4}(2-q)^3 & 1 \leqslant q < 2\\ 
        \\
        0 & q \geqslant 2
        \end{matrix}\right.
        $}
    \end{equation}

@@

    \begin{equation} \label{eq:gradKernel}
        \resizebox{0.88\linewidth}{!}{$
        \partial W_h(q(x)) =  x \frac{1}{h \left \| x \right \|} \frac{1}{\pi h^3} \left\{\begin{matrix}
        - 3q + \frac{9}{4}q^2 & 0 \leqslant q < 1 \\ 
        \\
        - \frac{3}{4}(2-q)^2 & 1 \leqslant q < 2\\ 
        \\
        0 & q \geqslant 2
        \end{matrix}\right.
        $}
    \end{equation}
    % maybe specify x here as well

@@

\subsection{Navier-Stokes}
    Navier-Stokes equations describes incompressible fluids and is therefore suitable for SPH simulations. The equations expressed in Lagrangian coordinates can be found in Equation~\ref{eq:nsDensity} and~\ref{eq:nsVelocity}.

    \begin{equation} \label{eq:nsDensity}
        \frac{D\rho}{Dt} = 0 \Leftrightarrow \bigtriangledown \cdot \textbf{v} = 0 
    \end{equation}

    \begin{equation} \label{eq:nsVelocity}
        \frac{D\textbf{v}}{Dt} = \frac{1}{\rho} \bigtriangledown p + v \bigtriangledown ^2 \textbf{v} + \frac{\textbf{f}}{\rho}
    \end{equation}

    In the equations $\frac{D\rho}{Dt}$ and $\frac{D\textbf{v}}{Dt}$ denotes the derivate of the density $\rho$ and the velocity $\textbf{v}$ respectively. The pressure, kinematic viscosity and body forces are denoted by $p,v$ and $\textbf{f}$ respectively. The second part of Equation~\ref{eq:nsDensity} $\bigtriangledown \cdot \textbf{v} = 0$ induce that the velocity field is free from divergence. From the continuity equation $\frac{D\rho}{Dt} = -\rho \bigtriangledown \cdot \textbf{v}$ and the divergence-free condition it follows that the partial derivate of the density with respect to the time is zero, which implies the equivalence of the equations in Equation~\ref{eq:nsDensity}. From this is follows that the density, in theory, must stay constant over time which is the implication of an incompressible fluid. In practice the divergence-free condition is not sufficient to guarantee incompressibility in simulations. The numerical time integrations will contain numerical errors which will cause density deviations due to the volume compressions. To avoid this a second condition $\rho - \rho_0 = 0$ called \textit{constant density condition} must be fullfilled~\cite{bender}.

\subsection{Divergence solver}
\subsection{Density solver}

% Write about non pressure forces, velocity osv..?
\subsection{Adapted time step}

% CFL condition, explain why its necessary and how its assured that each rendered frame has the same time step.
In order to assure a stable simulation it is necessary to adapt the time step depending on the motion. 
The quicker the particles move the smaller time steps are taken such that the numerical 
error of the explicit time integration is negligible. For this is a condition proposed by Courant–Friedrichs–Lewy used often refered as the CFL-condition show in equation \ref{eq:cfl}. The time step $\Delta t$ is equal to a fraction of the particle's diameter $d$ divided by the maximum velocity $\vect{V_{max}}$ in the scene.

@@

\subsection{Density and alpha factors}
\subsection{Screen space fluid rendering}

>>>>

\subsection{Neighbourhood search}
    Since SPH only considers a finite amount of neighbouring particles it is important to keep track of every particles neighbours.
    Searching through all particles for neighbours within the cutoff distance $H$ for every particle is inefficient and takes $\mathcal{O}({N^2})$ time.
    The cutoff distance $H$ is the kernels smoothing radius.
    To make this faster a cell list was implemented.
    A cell list is a data structure that is divided into cells that have a length larger or equals to the cutoff distance $H$.
    Each cell spans on a surface in space.
    A particle belongs to a cell if it is inside the space the cell occupy.
    When finding the neighbour of particle $i$, only the neighbouring cells have to be searched for particles within the cutoff distance, see figure~\ref{fig:cellList}.

    \begin{figure}[h!]\centering
    \begin{centering}
    \includegraphics[width=0.7\linewidth]{img/CellList.png}
    \caption{Finding the neighbours for the filled in particle $i$ by looking through all neighbouring cells, including its own cell, for particles within the cutoff distance $H$}
    \label{fig:cellList}
    \end{centering}
    \end{figure}

    The cell list is implemented by using a four dimensional vector where the first three dimensions are for the x, y and z coordinates and the fourth dimension is for storing the particles belonging to the cell.
    The amount of cells are decided by dividing the scene into cells of length $H$.
    The particles are then assigned to a cell by finding the cell's coordinates that it belongs to according to Equation ~\ref{eq:assignCell}, where $\vect{x}$ is the position of the particle and $\vect{s}_min$ is the lowest position of the scene space.
    If the particle moves out of its cell it is reassigned to the new cell it is inside.

    \begin{equation} \label{eq:assignCell}
        \frac{\lceil{\vect{x} - \vect{s}_{min}}\rceil}{H}
    \end{equation}

@@

    \begin{equation} \label{eq:kernel}
        W_h(q(x)) =  \frac{1}{\pi h^3} \left\{\begin{matrix}
        1 - \frac{3}{2}q^2 + \frac{3}{4}q^3 & 0 \leqslant q < 1 \\ 
        \\
        \frac{1}{4}(2-q)^3 & 1 \leqslant q < 2\\ 
        \\
        0 & q \geqslant 2
      \end{matrix}\right.
    \end{equation}

@@

    \begin{equation} \label{eq:gradKernel}
        \partial W_h(q(x)) =  x \cdot \frac{1}{h \left \| x \right \|} \cdot \frac{1}{\pi h^3} \left\{\begin{matrix}
        - 3q + \frac{9}{4}q^2 & 0 \leqslant q < 1 \\ 
        \\
        - \frac{3}{4}(2-q)^2 & 1 \leqslant q < 2\\ 
        \\
        0 & q \geqslant 2
      \end{matrix}\right.
    \end{equation}
    % maybe specify x here as well

@@

\subsection{Divergence-Free solver}

\iffalse
\begin{lstlisting}[language=C++, mathescape=true, escapeinside={(*}{*)}]
function correctDivergenceError
    for all particles $i$ do
        calculateDensityChange $\frac{Dp}{Dt}_i$
    for all particles $i$ do //Scale $\color{codegreen}a_i$
        $a_i = \frac{a_i}{\Delta t}$
    while (t < $t_{max}$) do //Start simulation loop
        adapt time step $\Delta{t}$ 
        for all particles $i$ do //Predict velocities $\color{codegreen}\mathbf{v}_i^*$
            $\mathbf{v}_i = \mathbf{v}_i + \Delta{t}\mathbf{F}_i^{adv}/m_i$
        correctDensityError($\alpha, \mathbf{v}_i$) //Fulfill $\color{codegreen}\rho^* - \rho_0 = 0$
        for all particles $i$ do //Update positions
            $x_i(t + \Delta{t}) = x_i(t) + \Delta{t}\mathbf{v}_i^*$
        for all particles $i$ do //Update neighbourhoods
            find neighbourhoods $N_i(t + \Delta{t})$
        for all particles $i$ do //Update $\color{codegreen}\rho_i$ and $\color{codegreen}\alpha_i$ 
            compute densities $\rho_i(0)$
            compute factors $\alpha_i(0)$
        correctDivergenceError($\alpha, \mathbf{v}^*$) //Fulfill $\color{codegreen}\frac{D\rho}{Dt} = 0$
\end{lstlisting}
\fi

The divergence-free solver ensures the fluid to be free of density change over time by enforcing $\frac{D\rho}{Dt} = 0$, which is equivalent to the Navier-Stokes divergence free condition given in equation %\ref{eq:navier}.

The solver will ensure that the divergence error is of an acceptable level defined by the user. The divergence error is suppressed by a combined symmetric pressure force $\vect{F_{j\leftarrow i}^p} $ over all particle $i$'s neighbours $j$, that is applied to the predicted velocity of particle $i$. The pressure force is determined by

\begin{equation} \label{eq:pressure}
\vect{F_i^p} = -\frac{m_i}{\rho_i}\nabla \vect{p_i},
\end{equation}

where the pressure gradient can be solved using the SPH formulation proposed by Ihmsen et al.~\cite{ihmsen}.

\begin{equation} \label{eq:dpdt}
\frac{D\rho_i}{Dt} = \sum_j m_j (\vect{v_i} - \vect{v_j})\nabla \vect{W_{ij}}
\end{equation}

\begin{equation} \label{eq:kappa}
\nabla \vect{p_i} = \kappa_i^v \nabla \rho_i = \kappa_i^v \sum_j m_j \nabla \vect{W_{ij}}
\end{equation}

Equation \ref{eq:dpdt} defines how the density change is computed and \ref{eq:kappa} computes the pressure gradient with the help of the stiffness parameter $\kappa_i^v$, where

\begin{equation}
\kappa_i^v = \frac{D\rho}{Dt} \cdot \frac{\rho_i}{\Delta t \cdot \alpha_i}
\end{equation}

and $\alpha_i$ is the scaling factor only dependant on the position explained in section \ref{section:alpha}. The stiffness parameters needs to be determined iteratively since neighboring particles depend on each other.
The velocity of particle $i$ is thus changed to:

\begin{equation} \label{eq:pressuretotal}
\vect{v_i^*}(t) = \vect{v_i^*}(t) - \Delta t \sum_j m_j (\frac{\kappa_i^v}{\rho_i} \cdot \frac{\kappa_j^v}{\rho_i}) \nabla \vect{W_{ij}}
\end{equation}

After the symmetric pressure force has been applied to the new velocity should the density change in equation \ref{eq:dpdt} be equal or close to zero. If the error is too large will the process be iterated until desired result have been achieved.

\subsection{Density solver}

\subsection{Navier-stokes}
% Write about non pressure forces, velocity osv..?
\subsection{Adapted time step}

% CFL condition, explain why its necessary and how its assured that each rendered frame has the same time step.
In order to assure a stable simulation it is necessary to adapt the time step depending on the motion. 
The quicker the particles move the smaller time steps are taken such that the numerical 
error of the explicit time integration is negligible. For this is a condition proposed by 
Courant$-$Friedrichs$-$Lewy used often refered as the CFL-condition show in equation \ref{eq:cfl}. 
The time step $\Delta t$ is equal to a fraction of the particle's diameter $d$ divided by 
the maximum velocity $\vect{V_{max}}$ in the scene.

@@

\subsection{Density and alpha factors} \label{section:alpha}
\subsection{Screen space fluid rendering}
